Decompositions of relative Chevalley groups

### 2015-10-07 paper.tex
Rewriting introduction

\begin{abstract} 
We formulate and prove relative versions of several classical decompositions known in the theory of Chevalley groups over commutative rings. % Gauss, Bass---Kolster and Dennis---Vaserstein decompositions.
As an application we obtain upper estimates for the width of principal congruence subgroups in terms of several families of generators.
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
%Let $R$ be a commutative unital ring and $\Phi$ be a reduced irreducible root system of rank $\ell$ with a fixed basis of simple roots.
Let $\G(\Phi, R)$ the split simply-connected Chevalley group of type $\Phi$ over arbitarary commutative ring $R$ and let $\E(\Phi, R)$ be its \emph{elementary subgroup} (see e.\,g.~\cite{St78, VP, S}).
%i.\,e. the subgroup generated by root unipotents $x_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in R$ see~\cite{VP, St78, S}.
For an ideal $I \trianglelefteq R$ we denote by $\G(\Phi, R, I)$ the \emph{relative Chevalley group} (aka \emph{principal congruence subgroup}) of level $I$, 
i.\,e. the kernel of the map $\G(\Phi, R)\rightarrow\G(\Phi, R/I)$ induced by the canonical projection $R\rightarrow R/I$.

The main goal of the present paper is to obtain relative analogues of several classical parabolic factorizations of Chevalley groups (i.\,e. factorizations into a product of several parabolic and unipotent subgroups)
%To be more specific, we study relative analogues of the following three kinds of decompositions:
%\begin{itemize}
%\item \emph{Gauss decomposition,} i.\,e. a decomposition of the form $\E(\Phi,R)=\B \cdot \U^- \cdot \U$, where $\B$ denotes the standard Borel subgroup,
%      and $\U$, $\U^-$ are the unipotent radicals of $\B$ and the opposite Borel subgroup $\B^-$;
%\item \emph{Bass---Kolster decomposition} $G = \Par_i \cdot \U_i^- \cdot \U_i \cdot \U_i^-$, where 
%      $\Par_i$, $\Par_i^-$ are the opposite maximal parabolic subgroups of $\G$ corresponding to $i$-th simple root of $\Phi$ and $\U_i$, $\U_i^-$ are their unipotent radicals;
%\item \emph{Dennis---Vaserstein decomposition} $\E(\Phi, R) = \EP_i \cdot (\U_i^- \cap \U_j^-) \cdot \EP_j$, $i\neq j$,
%      where $\EP_i$ denotes the subgroup generated by root unipotents $x_\alpha(\xi)$ contained in $\Par_i$.
%\end{itemize}
%TODO: Insert some links to VIP's papers :)

@@

Our first result is a relative version Gauss decomposition.
All the notation for the subgroups appearing in this theorem and the other theorems stated below is explained in~\ref{sec:prelim}.
\begin{thm}\label{thm:srRI1}
Let $\Phi$ be a root system, let $I$ be an ideal of arbitrary commutative ring $R$ such that $\sr(I)=1$.
Then the relative elementary Chevalley $\E(\Phi, R, I)$ group admits the following decomposition
\[ \E(\Phi,R,I) = \Hh(\Phi,R,I) \cdot \U(\Phi,I) \cdot \U(\Phi^-,I) \cdot \U(\Phi,I). \]
\end{thm}
The special case $I=R$ of the \cref{thm:srRI1} was obtained by the second named author in~\cite{Sm12}.
We also note that in special case of a semilocal ring $R$ the above theorem was proved by E.~Abe and K.~Suzuki, see~\cite[Proposition~2.3]{AbeSuzNormalSubgroups}.

Our next main result is a relative Dennis---Vaserstein decomposition which holds for a broader class of groups of rings as compared to \cref{thm:srRI1}.
\begin{thm}\label{thm:DennisVaserstein}
Let $\Phi \neq \rE_8$ be a reduced irreducible root system and let $\{ r, s \}$ be a pair of distinct terminal vertices of the Dynkin diagram of $\Phi$ the distance between which equals $d$.
Assume that $\Phi$ and $I$ satisfy either of the following assumptions:
\begin{enumerate}
 \item $\asr(I) \leq d$ for $\Phi=\rF_4$ or $\Phi=\rE_\ell$, $\{r, s\} = \{1, \ell \}$,
 \item $\sr(I) \leq d$, otherwise.
\end{enumerate}
Then $\E(\Phi, R, I)$ factors as $\EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s) \cdot \EP_s(R, I).$ \end{thm}
Even in the special case $I=R$ the above theorem contains several new cases as compared to M.~Stein's Theorems~2.5, 4.1, see~\cite{St78}.
The term ``Dennis---Vaserstein decomposition'' is due to A.~Suslin and M.~Tulenbaev (cf.~\cite[Lemma~2.1]{ST76}).

The condition $\asr(I)$ used in the statement of \cref{thm:DennisVaserstein} is a relative version of the \emph{absolute stable range} of D.~Estes and J.~Ohm which does not depend on the ambient ring $R$ (see section~\ref{sec:rel-asr}).

@@

On the other hand, much less is known about the width of the relative group $\G(\Phi, R, I)$ and the relative elementary group $\E(\Phi, R, I)$.
Recall that the \emph{relative elementary group} $\E(\Phi, R, I)$ is defined as the normal closure of $\E(\Phi, I)$ in $\E(\Phi, R)$, where
$\E(\Phi, I)$ denotes the subgroup of $\G_{sc}(\Phi, R)$ generated by all elementary root unipotents $x_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in I$, i.\,e. $\E(\Phi, R, I) = \E(\Phi, I)^{\E(\Phi, R)}$.

In fact, one can choose a smaller set of generators for $\E(\Phi, R, I)$. 
Let $\Sigma$ be any special subset of $\Phi$, (e.\,g. the set of simple roots $\Pi$ or the special part of $\Sigma_S$ of some parabolic subset of roots $S$, see section~\ref{sec:rootsys}).
Set $$\mathcal{X} = \{ x_\alpha(s) \mid \alpha\in \Phi,\ s\in I\};\ \ \mathcal{Z}(\Sigma) = \{ z_\alpha(s, \xi) \mid s \in I,\ \xi \in R, \alpha\in \Sigma \} \cup \mathcal{X}.$$

Now we are ready to state a recent result of A.~Stepanov.  
\begin{prop}[{\cite[Theorem~3.4]{S}}] \label{prop:Stepanov}
Let $\Phi$ be a reduced irreducible root system of rank $\geq 2$ and let $S \subset \Phi$ be any parabolic subset of roots.
Then $\E(\Phi, R, I)$ is generated as an abstract group by $\mathcal{Z}(\Sigma_S)$.
\end{prop}

>>>>

\begin{abstract} 
We formulate and prove relative versions of several classical decompositions known in the theory of Chevalley groups over commutative rings. % Gauss, Bass---Kolster and Dennis---Vaserstein decompositions.
As an application we obtain upper estimates for the width of principal congruence subgroups in terms of several families of generators.
Some of our results are new even in the absolute case and were previously studied only for groups over finite fields.
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
%Let $R$ be a commutative unital ring and $\Phi$ be a reduced irreducible root system of rank $\ell$ with a fixed basis of simple roots.
Let $\G(\Phi, R)$ the split simply-connected Chevalley group of type $\Phi$ over arbitarary commutative ring $R$ and let $\E(\Phi, R)$ be its \emph{elementary subgroup}
i.\,e. the subgroup generated by root unipotents $x_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in R$ see~\cite{VP, St78, S}.
For an ideal $I \trianglelefteq R$ we denote by $\G(\Phi, R, I)$ the \emph{relative Chevalley group} (aka \emph{principal congruence subgroup}) of level $I$, 
i.\,e. the kernel of the map $\G(\Phi, R)\rightarrow\G(\Phi, R/I)$ induced by the canonical projection $R\rightarrow R/I$.

In our theorems we also use the notion of the \emph{relative elementary Chevalley group} $\E(\Phi, R, I)$.
Recall that $\E(\Phi, R, I)$ is defined as the normal closure of the subgroup generated by the set $\mathcal{X} = \{x_\alpha(\xi) \mid \alpha\in \Phi,\ \xi \in I \}$.

The following recent result of A.~Stepanov served as a starting point for the present paper.
All the notation and definitions used below are explained properly in section~\ref{sec:prelim}.
\begin{prop}[{\cite[Theorem~3.4]{S}}] \label{prop:Stepanov}
Let $\Phi$ be a reduced irreducible root system of rank $\geq 2$, let $S \subset \Phi$ be any parabolic subset of roots.
Let $\Sigma_S$ be the special part of $S$, then $\E(\Phi, R, I)$ is generated as an abstract group by the following set of generators
$$\mathcal{Z}(\Sigma) = \mathcal{X} \cup \{ z_\alpha(s, \xi) \mid s \in I,\ \xi \in R, \alpha\in \Sigma \}.$$ \end{prop}

The first goal of the present paper is to obtain relative analogues of several classical parabolic factorizations of Chevalley groups 
(i.\,e. factorizations into a product of several parabolic and unipotent subgroups)
%To be more specific, we study relative analogues of the following three kinds of decompositions:
%\begin{itemize}
%\item \emph{Gauss decomposition,} i.\,e. a decomposition of the form $\E(\Phi,R)=\B \cdot \U^- \cdot \U$, where $\B$ denotes the standard Borel subgroup,
%      and $\U$, $\U^-$ are the unipotent radicals of $\B$ and the opposite Borel subgroup $\B^-$;
%\item \emph{Bass---Kolster decomposition} $G = \Par_i \cdot \U_i^- \cdot \U_i \cdot \U_i^-$, where 
%      $\Par_i$, $\Par_i^-$ are the opposite maximal parabolic subgroups of $\G$ corresponding to $i$-th simple root of $\Phi$ and $\U_i$, $\U_i^-$ are their unipotent radicals;
%\item \emph{Dennis---Vaserstein decomposition} $\E(\Phi, R) = \EP_i \cdot (\U_i^- \cap \U_j^-) \cdot \EP_j$, $i\neq j$,
%      where $\EP_i$ denotes the subgroup generated by root unipotents $x_\alpha(\xi)$ contained in $\Par_i$.
%\end{itemize}
%TODO: Insert some links to VIP's papers :)

@@

Our first result is a relative version of Gauss decomposition formulated in terms of the \emph{stable rank} of the ideal $I$ (see section~\ref{sec:stability-conditions} for the definition)
\begin{thm}\label{thm:srRI1}
Let $\Phi$ be a root system, let $I$ be an ideal of arbitrary commutative ring $R$ such that $\sr(I)=1$.
Then the relative elementary Chevalley $\E(\Phi, R, I)$ group admits the following decomposition
\[ \E(\Phi,R,I) = \Hh(\Phi,R,I) \cdot \U(\Phi,I) \cdot \U(\Phi^-,I) \cdot \U(\Phi,I). \]
\end{thm}
In the special case when $I$ is contained in the Jacobson radical of $R$ the above theorem boils down to~\cite[Proposition~2.3]{AbeSuzNormalSubgroups}. 
Notice that the special case $I=R$ was treated by the second named author in~\cite{Sm12}.

Our next main result is a relative Dennis---Vaserstein decomposition which holds for a broader class of groups of rings as compared to \cref{thm:srRI1}.
\begin{thm}\label{thm:DennisVaserstein}
Let $\Phi \neq \rE_8$ be a reduced irreducible root system and let $\{ r, s \}$ be a pair of distinct terminal vertices of the Dynkin diagram of $\Phi$ the distance between which equals $d$.
Assume that $\Phi$ and $I$ satisfy either of the following assumptions:
\begin{enumerate}
 \item $\asr(I) \leq d$ for $\Phi=\rF_4$ or $\Phi=\rE_\ell$, $\{r, s\} = \{1, \ell \}$,
 \item $\sr(I) \leq d$, otherwise.
\end{enumerate}
Then $\E(\Phi, R, I) = \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s) \cdot \EP_s(R, I).$ \end{thm}
Even in the special case $I=R$ the above theorem contains several new cases as compared to~\cite[Theorems~2.5, 4.1]{St78}.
The term ``Dennis---Vaserstein decomposition'' is due to A.~Suslin and M.~Tulenbaev (cf.~\cite[Lemma~2.1]{ST76}).

The condition $\asr(I)$ used in the statement of \cref{thm:DennisVaserstein} is a relative version of the \emph{absolute stable rank} of D.~Estes and J.~Ohm which does not depend on the ambient ring $R$ (see section~\ref{sec:rel-asr}).

@@

On the other hand, much less is known about the width of the relative group $\G(\Phi, R, I)$ and the relative elementary group $\E(\Phi, R, I)$.
